### 2022-11-28 intro.Rnw
Maria feedback

How can we know if a finding from a study is really true? For example, how can
we know if the protective effect of a vaccine found in a study is real? The
answer to this question is of highest importance to scientists and decision
makers, but typically we can never know for sure as any study result comes with
uncertainty.

@@

One way to come a little closer to the truth, however, is to repeat that
original study with new subjects. Such a \emph{replication study} may then yield
similar results which would make us more confident about the original finding,
or it may yield conflicting results which would lower our confidence.
Replication studies are thus an essential part of the scientific process as they
provide a means for substantiating genuine research findings and refuting
research findings which occurred merely by chance. For this reason,
``successful'' replication is often a requirement, for example, for acceptance
of newly proposed scientific theories (\eg{} a new physical model) or the
implementation of policies based on scientific knowledge (\eg{} market approval
of a drug). The results from replication studies may thus have real world
consequences, such as deciding whether we should get vaccinated (a consequence
that anyone who experienced the COVID-19 pandemic is very aware of).

Despite their tremendous importance, the traditional academic system has made it
unattractive for researchers to conduct replication studies; the currencies of
science --publications, citations, and grant money-- are typically easier to
acquire by conducting novel research. This is because until relatively recently,
top journals in many fields refused to publish replication studies as they were
considered not ``innovative'' enough. To build a successful career, researchers
thus often had no other choice than to concentrate their efforts on producing
novel and eye-catching research results.

@@

Most of these large-scale replication projects confirmed what many researchers
had feared; carefully conducted replication studies often show less impressive
results than their original counterparts, and the replicability of research
findings is surprisingly low on average. This realization led many to declare
science as being in a ``replication crisis''. Debates arose about whether or not
the crisis really existed, and who or what was to blame \citep{Gilbert2016,
  Amrhein2019}. Even the popular press became interested
\citep[\eg{}][]{Carey2015, Kovic2016, Achenbach2017, Devlin2018}, so that also
in the eyes of the public the credibility of science became seriously
threatened.

\begin{figure}[!htb]
<< "google-scholar-graph", fig.height = 2.5 >>=
## Data gathered on October 18 2022
searchTerm <- "Replication Study"
searchDate <- as.Date(x = "2022-10-18", format = "%Y-%m-%d", tz = "GMT")
searchDateFormat <- format(searchDate, format = "%B %d %Y")
resultsAnytime <- 114000
scholarDF <- data.frame(year = seq(2022, 1990, -1),
                        results = c(4530, 11400, 13600, 12200, 11300, 10700,
                                    9880, 9220, 8660, 8200, 7500, 6700, 6040,
                                    5370, 4680, 3890, 3380, 3070, 2790, 2520,
                                    2310, 2190, 2120, 2020, 1920, 1840, 1730,
                                    1490, 1270, 1230, 1140, 1070, 970))

## plot data
ggplot(data = filter(scholarDF, year < 2022),
       aes(x = year, y = results)) +
    geom_segment(aes(xend = year, yend = 0), alpha = 0.2, size = 0.4) +
    geom_point() +
    scale_y_continuous(breaks = seq(0, 15000, 2500)) +
    scale_x_continuous(breaks = seq(1990, 2020, 5)) +
    ## scale_y_log10() +
    labs(x = "Year", y = paste0('Results for "', searchTerm, '"')) +
    theme_bw() +
    theme(panel.grid.minor = element_blank(),
          panel.grid.major.x = element_blank()## ,
          ## panel.grid.major.y = element_line(colour = "grey", linetype = "dashed")
          )
@
\caption{Number of results per year for search term ``\Sexpr{searchTerm}'' on
  Google Scholar. The search was conducted on 18 October 2022.}
\label{fig:scholar}
\end{figure}

In the midst of this doomsday, % , particularly from the hard-hit
% field of psychology,
various reforms were implemented to safe the reputation of science and to
prevent a second crisis from happening \citep[for an overview see
\eg{}][]{Munafo2017}. For instance, many journals adopted the ``registered
report'' format \citep{Chambers2021} in which a study proposal is peer reviewed
\emph{before} the study is conducted, and which, if accepted, gives provisional
publication acceptance regardless of the study outcome. Similarly, digital
infrastructure platforms, such as zenodo (\url{https://zenodo.org/}) or the open
science framework (\url{https://osf.io/}), were created to facilitate
preregistration, preprints, code, and data sharing, all of which have
substantially increased over the last decade \citep{Kidwell2016, Nosek2018,
  Rawlinson2019}. The practice of conducting replication studies has also gained
popularity, and several journals and funders are now explicitly promoting and
funding replication research \citep{NWO2016, NSL2018, Nature2022}.
Figure~\ref{fig:scholar} illustrates this trend via the yearly number of results
for ``\Sexpr{searchTerm}'' on Google Scholar over the last three decades. We see
that the numbers have been rapidly growing, especially after the mid 2000s (with
a minor drop in 2021, perhaps because of research slowing down due to the
COVID-19 pandemic).

\subsection{The statistics of replication studies}
Despite the increased interest in replication studies, the research community
has not yet agreed on one important question: When is a replication study
successful? For this reason, replication researchers typically report the
results from different methods for assessing replication success. For example,
\citet{Opensc2015} state ``[t]here is no single standard for evaluating
replication success. We evaluated [replicability] using significance and P
values, effect sizes, subjective assessments of replication teams, and
meta-analyses of effect sizes'' (p. 11). In the following, I will give an
overview about these and other methods which have been used in practice.

\begin{table}[!htb]
  \centering
  \caption{Study level summary statistics for original and replication study.
    The cumulative distribution function of the standard normal distribution is
    denoted by $\Phi(\cdot)$, and the $1 - \alpha$ quantile of the standard
    normal distribution is denoted by $\Phi^{-1}(1 - \alpha) = \z{\alpha}$.
    Confidence intervals and $p$-values are based on a normal approximation.}
  \label{tab:meta}
  \begin{tabular}{lcc}
    \toprule
    & Original study & Replication study \\
    \midrule
    effect estimate & $\that_o$ & $\that_{r}$ \\
    standard error & $\sigma_{o}$ & $\sigma_{r}$ \\
    $(1 - \alpha)$ confidence interval &
    $[\that_{o} \pm \z{\alpha/2} \sigma_{o}]$ &
    $[\that_{r} \pm \z{\alpha/2} \sigma_{r}]$ \\
    % $[\that_{o} - \z{\alpha/2} \sigma_{o}, \that_{o} + \z{\alpha/2} \sigma_{o}]$ &
    % $[\that_{r} - \z{\alpha/2} \sigma_{r}, \that_{r} + \z{\alpha/2} \sigma_{r}]$ \\
    $z$-value & $z_{o} = \that_{o}/\sigma_{o}$ & $z_{r} = \that_{r}/\sigma_{r}$ \\
    $p$-value (two-sided) & $p_{o} = 2\{1 - \Phi(|z_{o}|)\}$ & $p_{r} = 2\{1 - \Phi(|z_{r}|)\}$ \\
    % One-sided $p$-value & $p_{o} = 1 - \Phi(|z_{o}|)$ & $p_{r} = 2\{1 - \Phi(|z_{r}|)\}$ \\
    \bottomrule
  \end{tabular}
\end{table}

Most methods for analyzing replication studies can be formulated in terms of
study level summary statistics as shown in Table~\ref{tab:meta}. % Effect
% estimates, standard error, confidence intervals, and $p$-values
All of these are routinely reported in researcher articles, and if one of them
is missing they typically can, under some assumptions, be back-calculated from
the others. Using summary statistics is also often the only possible way for
conducting the analysis as the raw data from the original study may not be
available to the replicators. The most important statistic is the effect
estimate $\that$. It provides an estimate of the underlying effect size $\theta$
which quantifies the effect or association of an intervention/exposure with an
outcome variable. Typical effect sizes are mean differences and correlations
(for continuous outcomes), odds ratios (for binary outcomes), or hazard ratios
(for time to event outcomes). Depending on the effect size type, a
transformation might be required so that the assumption of approximately
normally distributed effect estimates around the unknown effect size $\theta$
(for large enough sample sizes) is justifiable. This could be, for instance, the
Fisher $z$-transformation for correlations, or the log-transformation for
odds/hazard ratios \citep[chapter 11]{Cooper2019}. The associated standard error
$\sigma$ represents the statistical uncertainty of the estimate. Under the
assumption of (asymptotic) normality, confidence intervals for $\theta$ and
$p$-values for testing the null hypothesis $H_{0} \colon \theta = 0$ can be
computed as shown in Table~\ref{tab:meta}.

\begin{table}[!htb]
  \centering
  \caption{Statistical criteria for assessing replication success which have
    been used in practice \citep{Opensc2015, Camerer2016, Camerer2018, Cova2018,
      Errington2021}.}
  \label{tab:rsdef}
  \begin{tabular}{p{0.21\textwidth}p{0.7\textwidth}}
    \toprule
    Criterion type & Description \\
    \midrule
    Significance & The original and replication $p$-values are smaller than
                   a threshold $\alpha$ and their effect estimates show
                   the same direction
                   ($p_{o} < \alpha$, $p_{r} < \alpha$, and
                   $\sign(\that_{o}) = \sign(\that_{r})$). Usually $\alpha = 5\%$.
    \\
                   & \\
    Meta-analytic significance & The meta-analytic $p$-value is smaller than a threshold
                                 $\alpha$ ($p_{m} = 2(1 - \Phi(|\that_{m}|/\sigma_{m})) < \alpha$)
                                 where $\that_{m} =
                                 (\that_{o}/\sigma_{o}^{2} + \that_{r}/\sigma^{2}_{r})\sigma^{2}_{m}$
                                 is the  pooled effect estimate with standard error
                                 $\sigma_{m} = (1/\sigma^{2}_{o} + 1/\sigma^{2}_{r})^{-1/2}$.
                                 Usually $\alpha = 5\%$. \\
                   & \\
    Relative effect size & The effect estimate of the replication study goes in the same
                           direction and is at least as large as $d_{\text{min}}$ times
                           the original one ($d = \that_{r}/\that_{o} \geq d_{\text{min}}$).
                           Usually $d_{\text{min}} = 1$. \\
                   & \\
    Confidence interval & The replication effect estimate is contained in the
                          $(1 - \alpha)$ original confidence interval
                          ($\that_{r} \in [\that_{o} \pm z_{\alpha/2}\sigma_{o}]$).
                          Sometimes, also defined as the original effect estimate is
                          contained in the $(1 - \alpha)$ replication confidence interval
                          ($\that_{o} \in [\that_{r} \pm z_{\alpha/2}\sigma_{r}]$).
                          Usually $\alpha = 5\%$. \\
                   & \\
    Prediction interval ($Q$-test) & The replication effect estimate is contained with its
                                     $(1 - \alpha)$ prediction interval
                                     ($\that_{r} \in [\that_{o} \pm z_{\alpha/2}\sqrt{\sigma_{o}^{2}
                                     + \sigma_{r}^{2}}]$). Usually $\alpha = 5\%$.
                                     This criterion is equivalent to $p_{Q} \geq \alpha$ where
                                     $p_{Q}$ is the $p$-value from the meta-analytic $Q$-test.\footnotemark\\
    \bottomrule
  \end{tabular}
\end{table}
\footnotetext{The $p$-value from the $Q$-test is
  $p_{Q} = 2\{1 - \Phi(\surd Q)\}$ where
  $Q = \sum_{i \in \{o, r\}} (\that_{i} - \that_{m})^{2}\sigma_{i}^{-2} = (\that_{o}- \that_{r})^{2}(\sigma^{2}_{o} + \sigma^{2}_{r})^{-1}$
  is the $Q$-statistic and $\that_{m}$ is the fixed-effects pooled estimate as
  defined in Table~\ref{tab:rsdef}.}

Table~\ref{tab:rsdef} lists commonly used criteria for replication success in
terms of the summary statistics from Table~\ref{tab:meta}. The most popular
criterion defines replication success by simultaneous statistical significance
of original and replication study along with their effect estimates showing the
same direction. This approach is also called the ``two-trials rule''
\citep{Senn2008} in drug regulation or ``vote-counting'' in meta-analysis
\citep{Cooper2019}. The criterion can similarly be defined through one-sided
$p$-values, so that the original and replication effect estimate are not
required to show the same direction as this is taken into account by the
$p$-values. Some replication projects \citep{Opensc2015, Errington2021} also
defined replication success via simultaneous non-significance of original and
replication study ($p_{o} > \alpha$ and $p_{r} > \alpha$). However, with this
definition ``replication success'' can almost always be achieved by conducting
original and replication study with just very few samples so that the $p$-values
are large. The approach is also logically questionable as it could be seen as an
instance of the ``absence of evidence fallacy'' \citep{Altman1995} meaning that
the failure to find evidence against the null hypothesis is erroneously
interpreted as evidence for the null hypothesis. Meta-analytic extensions of the
significance approach define replication success by significance of a combined
effect estimate.\footnote{The meta-analytic approach can also be given a
  Bayesian interpretation: A flat prior distribution for the underlying effect
  size $\theta$ is updated by the data from original and replication study.
  Replication success defined by a meta-analytic $p$-value being smaller than
  $\alpha$ is then equivalent to replication success via a Bayesian posterior
  probability $\Pr(\theta > 0 \given \that_{o}, \that_{r}) > 1 - \alpha/2$,
  respectively, $\Pr(\theta < 0 \given \that_{o}, \that_{r}) > 1 - \alpha/2$
  depending on the direction of combined estimate.} Typically, the assumption of
a common underlying effect size is seen as reasonable so that fixed-effects
meta-analysis is used for pooling. Random-effects meta-analysis has mostly been
used if more than one replication study of the same original study are conducted
since in this case replicators are often interested in also understanding
between-replication heterogeneity \citep[\eg{} in][]{Klein2018}.

The remaining criteria in Table~\ref{tab:rsdef} put more emphasis on
compatibility in effect size between original and replication study. For
example, the relative effect estimate $d = \that_{r}/\that_{o}$ quantifies how
much the replication effect estimate changed compared to the original one, and
the smaller $d$ the smaller the degree of replication success. Some projects
also report a confidence interval for $d$ \citep{Camerer2016, Camerer2018},
while others ignore its uncertainty and make a binary cut at one to define
replication success \citep{Errington2021}.

In contrast, the criteria based on confidence and prediction intervals define
effect size compatibility on an absolute scale. However, the criterion based on
the original confidence interval ignores the uncertainty from the replication,
whereas the criterion based on the replication confidence interval ignores the
uncertainty from the original study. The prediction interval criterion, on the
other hand, takes into account both sources of uncertainty \citep{Patil2016}.
Yet, also declaring ``replication success'' based on the prediction interval may
be logically questionable due to its connection to the meta-analytic $Q$-test.
% as it may also be seen as an instance of the ``absence of evidence fallacy''
% \citep{Altman1995};
That is, if the $p$-value from the $Q$-test is larger than $\alpha$ this is equivalent
to the replication effect estimate $\that_{r}$ being contained in its
$(1- \alpha)$ prediction interval. The null hypothesis of this test is defined
that the underlying effect sizes of original and replication study are the same
($H_{0} \colon \theta_{o} = \theta_{r}$), so a rejection of this null hypothesis
corresponds to demonstrating replication failure and not replication success.
Interpreting a failure to reject the null hypothesis as evidence for it is again
an instance of the absence of evidence fallacy
\citep{Hedges2019}. % To establish replication success a
% test based on the null hypothesis $H_{0} \colon \theta_{o} \neq \theta_{r}$
% would be required \citep{Hedges2019}.
As in the case of defining replication success by simultaneous non-significance
of original and replication study, the mismatch of the null hypothesis of the
prediction interval criterion results in the undesirable property that
``replication success'' can almost always be achieved if the sample size of the
studies is small enough as the prediction interval becomes wider with larger
standard errors.

@@

\subsubsection{Example: Reproducibility Project Cancer Biology}
I will now illustrate the assessment of replicability on data from the
Reproducibility Project: Cancer Biology \citep{Errington2021}. This large-scale
project attempted to replicate 53 landmark studies from the field of cancer
biology. However, due to various difficulties (\eg{} missing information from
the original studies or problems in conducting the experiments) only 23 of the
53 studies could be repeated. \citet{Errington2021} report that these
experiments led to data on 158 quantitative effects. However, from the data
which they provide only 132 quantitative effects come with original and
replication standardized mean difference effect estimates along with standard
errors, and only these data will be used in the subsequent analyses.

@@

plotB <- ggplot(data = rpcbWithoutNA, aes(x = smdo, y = smdr)) +
    geom_abline(intercept = 0, slope = 1, alpha = 0.3) +
    geom_hline(yintercept = 0, lty = 2, alpha = 0.9, size = 0.4) +
    geom_vline(xintercept = 0, lty = 2, alpha = 0.9, size = 0.4) +
    geom_point(aes(fill = rsig), shape = 21, alpha = 0.8,
               show.legend = FALSE) +
    geom_rug(aes(color = rsig), size = 0.3, show.legend = FALSE) +
    labs(x = "Original effect estimate (SMD)",
         y = "",
         fill = "") +
    coord_fixed(xlim = zoom, ylim = zoom) +
    theme_bw() +
    theme(panel.grid = element_blank())
ggpubr::ggarrange(plotA, plotB, ncol = 2, align = "v")
@
\caption{Results for 132 effects from the Reproducibility Project: Cancer
  Biology \citep{Errington2021} for which effect estimates and standard errors
  are available on standardized mean difference scale. The right plot shows a
  zoomed-in view of the dotted area in the left plot. The \textit{p}-values are
  recomputed using a normal approximation. Two study pairs with original effect
  estimate larger than 80 are not shown.}
\label{fig:rpcb}
\end{figure}

@@

Taken together, this analysis demonstrates that conclusions based on commonly
used replication success criteria often differ. It is not always clear cut
whether or not a replication is successful. Reducing replicability to a single
criterion without mentioning these difficulties, as often done by the popular
press \citep[\eg{} ``more than half of the findings did not hold up when
retested'' in][]{Carey2015}, is a clear simplification of the matter.
% that may harm the credibility of science.

@@

\newpage
\section{Thesis summary}
\label{sec:summary}
\raggedbottom

>>>>

How can we know if a finding from a scientific study is really true? For
example, how can we know if the protective effect of a vaccine found in a study
is real? The answer to this question is of highest importance to scientists and
decision makers, but typically we can never know for sure as any study result
comes with uncertainty.

@@

One way to come a little closer to the truth, however, is to repeat that
original study with new subjects. Such a \emph{replication study} may yield
similar results which would make us more confident about the original finding,
or it may yield conflicting results which would lower our confidence.
Replication studies are thus an essential part of the scientific process as they
provide a means for substantiating genuine research findings and refuting
research findings which occurred merely by chance. For this reason,
``successful'' replication is often a requirement, for example, for acceptance
of newly proposed scientific theories (\eg{} a new physical model) or the
implementation of policies based on scientific knowledge (\eg{} market approval
of a drug). The results from replication studies may thus have real world
consequences, such as deciding whether we receive the recommendation to get
vaccinated (a consequence that anyone who experienced the COVID-19 pandemic is
very aware of).

Despite their tremendous importance, the traditional academic system has made it
unattractive for researchers to conduct replication studies; the currencies of
science -- publications, citations, and grant money -- are typically easier to
acquire by conducting novel research. This is because until relatively recently,
top journals in many fields refused to publish replication studies as they were
not considered ``innovative'' enough. To build a successful career, researchers
thus often had no other choice than to concentrate their efforts on producing
novel and eye-catching research results.

@@

Most of these large-scale replication projects confirmed what many researchers
had feared; carefully conducted replication studies often show less impressive
results than their original counterparts, and the replicability of research
findings is surprisingly low on average. This realization led many to declare
science as being in a ``replication crisis''. Debates arose about whether or not
the crisis really existed and who or what was to blame \citep{Gilbert2016,
  Amrhein2019}. Even the popular press became interested
\citep[\eg{}][]{Carey2015, Kovic2016, Achenbach2017, Devlin2018}, so that also
in the eyes of the public the credibility of science became seriously
threatened.

\begin{figure}[!htb]
<< "google-scholar-graph", fig.height = 2.5 >>=
## Data gathered on October 18 2022
searchTerm <- "Replication Study"
searchDate <- as.Date(x = "2022-10-18", format = "%Y-%m-%d", tz = "GMT")
searchDateFormat <- format(searchDate, format = "%B %d %Y")
resultsAnytime <- 114000
scholarDF <- data.frame(year = seq(2022, 1990, -1),
                        results = c(4530, 11400, 13600, 12200, 11300, 10700,
                                    9880, 9220, 8660, 8200, 7500, 6700, 6040,
                                    5370, 4680, 3890, 3380, 3070, 2790, 2520,
                                    2310, 2190, 2120, 2020, 1920, 1840, 1730,
                                    1490, 1270, 1230, 1140, 1070, 970),
                        study = c(156000, 487000, 1820000, 959000, 2360000,
                                  675000, 3000000, 688000, 3150000, 512000,
                                  3630000, 451000, 3650000, 396000, 3680000,
                                  389000, 3890000, 375000, 3950000, 336000,
                                  4160000, 364000, 4240000, 385000, 3470000,
                                  377000, 3120000, 335000, 2620000, 376000,
                                  2440000, 317000, 2220000))

## plot data
ggplot(data = filter(scholarDF, year < 2022),
       aes(x = year, y = results)) +
    geom_segment(aes(xend = year, yend = 0), alpha = 0.2, size = 0.4) +
    geom_point() +
    ## scale_y_continuous(breaks = seq(0, 15000, 2500)) +
    scale_x_continuous(breaks = seq(1990, 2020, 5)) +
    ## scale_y_log10() +
    labs(x = "Year", y = paste0('Results for "', searchTerm, '"')) +
    theme_bw() +
    theme(panel.grid.minor = element_blank(),
          panel.grid.major.x = element_blank()## ,
          ## panel.grid.major.y = element_line(colour = "grey", linetype = "dashed")
          )
@
\caption{Number of results per year for search term ``\Sexpr{searchTerm}'' on
  Google Scholar. The search was conducted on 18 October 2022.}
\label{fig:scholar}
\end{figure}

In the midst of this doomsday, % , particularly from the hard-hit
% field of psychology,
various reforms were implemented to safe the reputation of science and to
prevent an escalation of the situation \citep[for an overview see
\eg{}][]{Munafo2017}. For instance, many journals adopted the ``registered
report'' format \citep{Chambers2021} in which a study proposal is peer reviewed
\emph{before} the study is conducted, and which, if accepted, gives provisional
publication acceptance regardless of the study outcome. Similarly, digital
infrastructure platforms, such as zenodo (\url{https://zenodo.org/}) or the open
science framework (\url{https://osf.io/}), were created to facilitate
preregistration, preprints, and code and data sharing, all of which have
substantially increased over the last decade \citep{Kidwell2016, Nosek2018,
  Rawlinson2019}. The practice of conducting replication studies has also gained
popularity, and several academic journals and funders are now explicitly
promoting and funding replication research \citep{NWO2016, NSL2018, Nature2022}.
Figure~\ref{fig:scholar} illustrates this trend via the yearly number of results
for ``\Sexpr{searchTerm}'' on Google Scholar over the last three decades. We see
that the numbers have been rapidly growing, especially after the mid 2000s (with
a minor drop in 2021, perhaps because of research slowing down due to the
COVID-19 pandemic).

\subsection{The statistics of replication studies}
Despite the increased interest in replication studies, the research community
has not yet agreed on one important question: When is a replication study
considered successful? For this reason, replication researchers typically report
the results from different methods for assessing replication success. For
example, the \citet{Opensc2015} states ``[t]here is no single standard for
evaluating replication success. We evaluated [replicability] using significance
and P values, effect sizes, subjective assessments of replication teams, and
meta-analyses of effect sizes'' (p. 11). In the following, I will give an
overview about these and other methods which have been used in practice.

\begin{table}[!htb]
  \centering
  \caption{Study level summary statistics for an original and a replication
    study. The cumulative distribution function of the standard normal
    distribution is denoted by $\Phi(\cdot)$, and the $1 - \alpha$ quantile of
    the standard normal distribution is denoted by
    $\Phi^{-1}(1 - \alpha) = \z{\alpha}$. Confidence intervals and $p$-values
    are based on the normality assumption.}
  \label{tab:meta}
  \begin{tabular}{lcc}
    \toprule
    & Original study & Replication study \\
    \midrule
    effect estimate & $\that_o$ & $\that_{r}$ \\
    standard error & $\sigma_{o}$ & $\sigma_{r}$ \\
    $(1 - \alpha)$ confidence interval &
    $[\that_{o} \pm \z{\alpha/2} \sigma_{o}]$ &
    $[\that_{r} \pm \z{\alpha/2} \sigma_{r}]$ \\
    % $[\that_{o} - \z{\alpha/2} \sigma_{o}, \that_{o} + \z{\alpha/2} \sigma_{o}]$ &
    % $[\that_{r} - \z{\alpha/2} \sigma_{r}, \that_{r} + \z{\alpha/2} \sigma_{r}]$ \\
    $z$-value & $z_{o} = \that_{o}/\sigma_{o}$ & $z_{r} = \that_{r}/\sigma_{r}$ \\
    $p$-value (two-sided) & $p_{o} = 2\{1 - \Phi(|z_{o}|)\}$ & $p_{r} = 2\{1 - \Phi(|z_{r}|)\}$ \\
    % One-sided $p$-value & $p_{o} = 1 - \Phi(|z_{o}|)$ & $p_{r} = 2\{1 - \Phi(|z_{r}|)\}$ \\
    \bottomrule
  \end{tabular}
\end{table}

Most methods for analyzing replication studies can be formulated in terms of
study level summary statistics as shown in Table~\ref{tab:meta}. % Effect
% estimates, standard error, confidence intervals, and $p$-values
All of these are routinely reported in researcher articles, and if one of them
is missing they typically can, under some assumptions, be calculated from the
others. Using summary statistics is also often the only possible way for
conducting the analysis as the raw data from the original study may not be
available to the replicators. The most important statistic is the effect
estimate $\that$. It provides an estimate of the underlying effect size $\theta$
which quantifies the effect or association of an intervention/exposure with an
outcome variable. Typical effect sizes are mean differences and correlations
(for continuous outcomes), odds ratios (for binary outcomes), or hazard ratios
(for time to event outcomes). Depending on the type of effect size, a
transformation might be required so that the assumption of approximately
normally distributed effect estimates around the unknown effect size $\theta$
(for large enough sample sizes) is justified. This could be, for instance, the
Fisher $z$-transformation for correlations, or the log-transformation for
odds/hazard ratios \citep[chapter 11]{Cooper2019}. The associated standard error
$\sigma$ represents the statistical uncertainty of the estimate. Under the
assumption of (asymptotic) normality, confidence intervals for $\theta$ and
$p$-values for testing the null hypothesis $H_{0} \colon \theta = 0$ can be
computed as shown in Table~\ref{tab:meta}.

\begin{table}[!htb]
  \centering
  \caption{Statistical criteria for assessing replication success which have
    been used in practice \citep{Opensc2015, Camerer2016, Camerer2018, Cova2018,
      Errington2021}.}
  \label{tab:rsdef}
  \begin{tabular}{p{0.21\textwidth}p{0.7\textwidth}}
    \toprule
    Criterion type & Description \\
    \midrule
    Significance & The original and replication $p$-values are smaller than
                   a threshold $\alpha$ and their effect estimates show
                   the same direction
                   (\mbox{$p_{o} < \alpha$}, \mbox{$p_{r} < \alpha$}, and
                   \mbox{$\sign(\that_{o}) = \sign(\that_{r})$}). Usually $\alpha = 5\%$.
    \\
                   & \\
    Meta-analytic significance & The meta-analytic $p$-value is smaller than a threshold
                                 $\alpha$ (\mbox{$p_{m} = 2\{1 - \Phi(|\that_{m}|/\sigma_{m})\} < \alpha$})
                                 where \mbox{$\that_{m} =
                                 (\that_{o}/\sigma_{o}^{2} + \that_{r}/\sigma^{2}_{r})\sigma^{2}_{m}$}
                                 is the  pooled effect estimate with standard error
                                 \mbox{$\sigma_{m} = (1/\sigma^{2}_{o} + 1/\sigma^{2}_{r})^{-1/2}$}.
                                 Usually $\alpha = 5\%$. \\
                   & \\
    Relative effect size & The effect estimate of the replication study goes in the same
                           direction as the original one and is at least $d_{\text{min}}$ times as large (\mbox{$d = \that_{r}/\that_{o} \geq d_{\text{min}}$}).
                           Usually $d_{\text{min}} = 1$. \\
                   & \\
    Confidence interval & The replication effect estimate is contained in the
                          $(1 - \alpha)$ original confidence interval
                          ($\that_{r} \in [\that_{o} \pm z_{\alpha/2}\sigma_{o}]$).
                          Sometimes, also defined as the original effect estimate is
                          contained in the $(1 - \alpha)$ replication confidence interval
                          ($\that_{o} \in [\that_{r} \pm z_{\alpha/2}\sigma_{r}]$).
                          Usually $\alpha = 5\%$. \\
                   & \\
    Prediction interval ($Q$-test) & The replication effect estimate is contained with its
                                     $(1 - \alpha)$ prediction interval
                                     ($\that_{r} \in [\that_{o} \pm z_{\alpha/2}\surd(\sigma_{o}^{2}
                                     + \sigma_{r}^{2})]$). Usually $\alpha = 5\%$.
                                     This criterion is equivalent to $p_{Q} \geq \alpha$ where
                                     $p_{Q}$ is the $p$-value from the meta-analytic $Q$-test
                                     $p_{Q} = 2\{1 - \Phi(\surd Q)\}$ with $Q$-statistic
                                     $Q = \sum_{i \in \{o, r\}} (\that_{i} - \that_{m})^{2}/\sigma_{i}^{2} = (\that_{o}- \that_{r})^{2}/(\sigma^{2}_{o} + \sigma^{2}_{r})$.\\
    \bottomrule
  \end{tabular}
\end{table}

Table~\ref{tab:rsdef} lists commonly used criteria for replication success in
terms of the summary statistics from Table~\ref{tab:meta}. The most popular
criterion is listed first and defines replication success by simultaneous
statistical significance of original and replication study along with their
effect estimates showing the same direction. This approach is also called the
``two-trials rule'' \citep{Senn2008} in drug regulation or ``vote-counting'' in
meta-analysis \citep{Cooper2019}. The criterion can similarly be defined through
one-sided $p$-values, so that the original and replication effect estimate are
not required to show the same direction as this is taken into account by the
$p$-values. Some replication projects \citep{Opensc2015, Errington2021} also
defined replication success via simultaneous non-significance of original and
replication study ($p_{o} > \alpha$ and $p_{r} > \alpha$). However, with this
definition ``replication success'' can almost always be achieved by conducting
original and replication study with very few samples so that the $p$-values are
large. The approach is also logically questionable as it could be seen as an
instance of the ``absence of evidence fallacy'' \citep{Altman1995} meaning the
failure to find evidence against the null hypothesis is erroneously interpreted
as evidence for the null hypothesis. Meta-analytic extensions of the
significance approach define replication success by significance of a combined
effect estimate.\footnote{The meta-analytic approach can also be given a
  Bayesian interpretation: A flat prior distribution for the underlying effect
  size $\theta$ is updated by the data from original and replication study.
  Replication success defined by a meta-analytic $p$-value being smaller than
  $\alpha$ is then equivalent to replication success via a Bayesian posterior
  probability $\Pr(\theta > 0 \given \that_{o}, \that_{r}) > 1 - \alpha/2$,
  respectively, $\Pr(\theta < 0 \given \that_{o}, \that_{r}) > 1 - \alpha/2$
  depending on the direction of combined estimate.} Typically, the assumption of
a common underlying effect size is seen as reasonable so that fixed-effects
meta-analysis is used for pooling. Random-effects meta-analysis has primarily
been used if more than one replication study of the same original study are
conducted since in this case replicators are often interested in also
understanding between-replication heterogeneity \citep[\eg{} in][]{Klein2018}.

The remaining criteria in Table~\ref{tab:rsdef} put more emphasis on
compatibility in effect size between the original and the replication study. For
example, the relative effect estimate $d = \that_{r}/\that_{o}$ quantifies how
much the replication effect estimate changed compared to the original, and the
smaller $d$ is the smaller the degree of replication success. Some projects also
report a confidence interval for $d$ \citep{Camerer2016, Camerer2018}, while
others ignore its uncertainty and make a binary cut at $d_{\text{min}} = 1$ to
define replication success \citep{Errington2021}. In contrast, the criteria
based on confidence and prediction intervals define effect size compatibility on
an absolute scale. However, the criterion based on the original confidence
interval ignores the uncertainty from the replication, whereas the criterion
based on the replication confidence interval ignores the uncertainty from the
original study. The prediction interval criterion, on the other hand, takes into
account both sources of uncertainty \citep{Patil2016}. Yet, declaring
``replication success'' based on the prediction interval may be logically
questionable due to its connection to the meta-analytic $Q$-test.
% as it may also be seen as an instance of the ``absence of evidence fallacy''
% \citep{Altman1995};
That is, if the $p$-value from the $Q$-test is larger than $\alpha$ this is
equivalent to the replication effect estimate $\that_{r}$ being contained in its
$(1- \alpha)$ prediction interval. The null hypothesis of this test is defined
that the underlying effect sizes of original and replication study are the same
($H_{0} \colon \theta_{o} = \theta_{r}$), so a rejection of this null hypothesis
corresponds to demonstrating replication failure and not replication success.
Interpreting a failure to reject the null hypothesis as evidence for it is again
an instance of the ``absence of evidence fallacy''
\citep{Hedges2019}. % To establish replication success a
% test based on the null hypothesis $H_{0} \colon \theta_{o} \neq \theta_{r}$
% would be required \citep{Hedges2019}.
As in the case of defining replication success by simultaneous non-significance
of original and replication study, the mismatch of the null hypothesis of the
prediction interval criterion results in the undesirable property that
``replication success'' can almost always be achieved if the sample size of the
studies is small enough as the prediction interval becomes wider with larger
standard errors.

@@

\subsubsection{Example: Reproducibility Project Cancer Biology}
I will now illustrate the assessment of replicability on data from the
Reproducibility Project: Cancer Biology \citep{Errington2021}. This large-scale
project attempted to replicate 53 landmark studies from the field of cancer
biology. However, due to various difficulties (\eg{} missing information from
the original studies or problems in conducting the experiments) only 23 of the
53 studies could be repeated. \citet{Errington2021} report that these
experiments led to data on 158 quantitative effects. However, from the data
which they provide only 132 quantitative effects are provided with original and
replication standardized mean difference effect estimates along with standard
errors, and only these data will be used in the subsequent analyses.

@@

plotB <- ggplot(data = rpcbWithoutNA, aes(x = smdo, y = smdr)) +
    geom_abline(intercept = 0, slope = 1, alpha = 0.3) +
    geom_hline(yintercept = 0, lty = 2, alpha = 0.9, size = 0.4) +
    geom_vline(xintercept = 0, lty = 2, alpha = 0.9, size = 0.4) +
    geom_point(aes(fill = rsig), shape = 21, alpha = 0.8,
               show.legend = FALSE) +
    geom_rug(aes(color = rsig), size = 0.3, show.legend = FALSE) +
    labs(x = "Original effect estimate (SMD)",
         y = "Replication effect estimate (SMD)",
         fill = "") +
    coord_fixed(xlim = zoom, ylim = zoom) +
    theme_bw() +
    theme(panel.grid = element_blank())
ggpubr::ggarrange(plotA, plotB, ncol = 2, align = "v")
@
\caption{Results for 132 effects from the Reproducibility Project: Cancer
  Biology \citep{Errington2021} for which effect estimates and standard errors
  are available on standardized mean difference (SMD) scale. The right plot
  shows a zoomed-in view of the dotted area in the left plot. The
  \textit{p}-values are recomputed using a normal approximation. Two study pairs
  with original effect estimate larger than 80 are not shown.}
\label{fig:rpcb}
\end{figure}

@@

Taken together, this analysis demonstrates that conclusions based on commonly
used replication success criteria often differ. It is not always clear cut
whether or not a replication is successful. Reducing replicability to a single
criterion without mentioning these difficulties, as often done by the popular
press \citep[\eg{} ``more than half of the findings did not hold up when
retested'' in][]{Carey2015}, is a simplification of the matter.
% that may harm the credibility of science.

@@

\newpage
\section{Open questions and thesis contributions}
\label{sec:summary}
\raggedbottom
